Semantic Linearity in a Lazy Optimising Compiler

### 2024-01-30 icfp.tex
up to 5.2, now at 25 pages

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Metatheory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Metatheory\label{sec:main:metatheory}}

The $\lambda^\pi_\D$ system is sound: well-typed programs in Linear Core do not
get \emph{stuck}. Besides type safety ($\S$~\ref{sec:type-safety-meta}), we
prove multiple optimising transformations preserve linearity
($\S$~\ref{sec:optimisations-preserve-types-meta}), and prove an auxiliary
result regarding proof irrelevant resources, stating that a case alternative
well-typed in a proof irrelevant context is also well-typed if proof irrelevant
resources are substituted by an arbitrary environment of relevant resources.
%
Additionally, we state our assumptions that outline an isomorphism between
using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
resources $\D$, for any $\D$.

\subsection{Assumptions}

\renewcommand{\DeltaLinearRelationLemma}{
\begin{lemma}[$\D \Rightarrow 1$]
Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env. is updated accordingly).\\
If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
\end{lemma}
}

\renewcommand{\LinearDeltaRelationLemma}{
\begin{lemma}[$1 \Rightarrow \D$]
A linear variable can be moved to the unrestricted context as a $\D$-var with any usage environment $\D$ by introducing $\D$ in the linear resources (and the unrestricted env. is updated accordingly).\\
If $\G; \D',\xl \vdash e :\vp$
then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$.
\end{lemma}
}

\renewcommand{\DeltaUnrestrictedRelationLemma}{
\begin{lemma}[$x{:}_\omega\sigma = x{:}_{\cdot}\sigma$]
An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
\end{lemma}
}

We use two main assumptions in our proofs, which are dual.
%
First, a program well-typed with a linear variable ($\xl$) is equivalently
well-typed if that same linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, $\D$ were available in the linear context instead of
the linear variable.
% , and occurrences of $x$ were substituted by $\D$ in the
% usage environments of other $\D$-vars in $\G$.
%
% First, we state that a linear variable $\xl$ can be replaced by a $\D$-bound
% variable as long as the context
%
\LinearDeltaRelationLemma
%
\noindent Second, a program well-typed with resources $\D$ and $\D$-bound
variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
through the use of $x$, if $x$ is moved to the linear context, resources $\D$
are removed from the linear context, and occurrences of $\D$ whole in usage
environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
environments are unimportant since $\D$ was consumed whole by $x$, not by any
of the fragment-using $\D$-vars).

@@

We additionally state that unrestricted resources are equivalent to $\D$-bound
variables with an empty ($\cdot$) usage environment:

@@

\subsection{Irrelevance}

As discussed above, proof irrelevant resources are resources that can only be
consumed indirectly, and are used to type case expressions whose scrutinee is
not in WHNF, essentially encoding that the scrutinee resources must be consumed
through the case binder or the linear pattern-bound variables.
%
As a case expression is evaluated, the scrutinee will eventually be in WHNF,
which must then be typed with rule $Case_{\textrm{WHNF}}$.
%
Crucially, these rules must ``work together'' in the system, in the sense that
case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
well-typed after the scrutinee is evaluated to WHNF, which is then typed using
the $Case_{\textrm{WHNF}}$ rule.

@@

The \emph{Irrelevance} lemma witnesses the soundness of typing a
case alternative with proof irrelevant resources in a certain context with respect to
typing the same expression with arbitrary resources (we note, however, typing
an alternative with proof irrelevant resources is not complete wrt using
arbitrary resources -- a counter example needs only to use a resource
directly).

>>>>

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Metatheory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Metatheory of Linear Core\label{sec:main:metatheory}}

The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
get \emph{stuck}. Besides type safety, we
prove that multiple optimising transformations used in GHC preserve
linearity in the sense of being well-typed in our system.
%
% Additionally, we state our assumptions that outline an isomorphism between
% using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
% resources $\D$, for any $\D$.

\renewcommand{\DeltaLinearRelationLemma}{
\begin{lemma}[$\Delta$-bound to Linear]
% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env. is updated accordingly).\\
If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
\end{lemma}
}

\renewcommand{\LinearDeltaRelationLemma}{
  \begin{lemma}[Linear to $\Delta$-bound]
\\
If $\G; \D',\xl \vdash e :\vp$
then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$.
\end{lemma}
}

\renewcommand{\DeltaUnrestrictedRelationLemma}{
\begin{lemma}[Unrestricted and $\Delta$-bound]
%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
\end{lemma}
}

\paragraph{$\Delta$-bound variables} Our development relies on the following characterization of $\D$-bound
variables:
A well-typed program with a linear variable ($\xl$) is equivalently
well-typed if the linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, with $\D$ available in the linear context instead of
the linear variable (and the unrestricted context updated accordingly).
\LinearDeltaRelationLemma

\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
environment can be made into a linear variable as long as its
(irrelevant) environment is dropped from the linear context.

% Dually, a well-typed program with resources $\D$ and $\D$-bound
% variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
% through the use of $x$, if $x$ is moved to the linear context, resources $\D$
% are removed from the linear context, and occurrences of $\D$ whole in usage
% environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
% environments are unimportant since $\D$ was consumed whole by $x$, not by any
% of the fragment-using $\D$-vars).

@@

\noindent Finally, unrestricted resources are equivalent to $\D$-bound
variables with an empty ($\cdot$) usage environment.

@@

\paragraph{Irrelevance}
%
As discussed, proof irrelevant resources can only be consumed
indirectly, essentially encoding that the (non-WHNF) case scrutinee
resources must be consumed through the case binder or the linear
pattern-bound variables.
%
As a case expression is evaluated, the scrutinee will eventually
reduce to WHNF,
which must then be typed with rule $Case_{\textrm{WHNF}}$.
%
Crucially, the two case rules must be in harmony in the system, in the sense that
case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.

@@

The \emph{Irrelevance} lemma witnesses the soundness of typing a
case alternative with proof irrelevant resources with respect to
typing the same expression with arbitrary resources. However, typing
a case alternative with proof irrelevant resources is not complete wrt using
arbitrary resources -- a counter example needs only to use a resource
directly.
